---
title: "Deploying R to PL/R"
author: "William Schneider"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deploying R to PL/R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

As part of development on big data platforms, it is often the case that a 
data scientist needs to deploy a R script, which runs on a locally-developed 
memory-limited environment, on a database to scale. PL/R provides a way to do that, however, 
the steps to achieve this can be automated. This package provides a few 
functions which alleviate the manual editing of PL/R function definitions 
as well as streamline the development process after the algorithm is 
executed in-database.

## Development Workflow

Since R can only be executed dynamically on the developer's local machine, 
the developer can only 'see' into the internal data by pulling data from 
the database and running the R script. To do this in a way which mimicks the 
form in which data would be available in the PL/R function, the function 
`pullDataSet` executes parameterized SQL queries, which internally uses 
the `RPostgreSQL` library.

As the developer works, he/she must present results to others, prove / disprove 
hypotheses, perform unit-tests, etc., in addition to developing the 
algorithm. Using a 
R markdown file, all these activities can be performed in one document, 
as opposed to having a local-machine-only version which has statetments not 
intended for execution on the database.

The next function in the package, `compile`, takes the working document and 
turns it into a true PL/R function. Additionally, it parses out the 
working material which isn't intended to be executed on the database. It 
uses the R markdown syntax to parse, and names for removing content not 
needed.

## Example

Suppose we have a dataset of a graph with nodes of various classes in a 
database called `graphs`. We are 
interested in applying a model to nodes of a given class. We can therefore 
create a query such as `SELECT * FROM graph WHERE class = v_class` 
and store it in a .sql file, where the `v_class` is a parameter where we 
can select a certain class. We would like to use this query in the R 
environment with a R markdown file.

The first thing we need in the file is the connection to the database:

```{r prepConn, eval= FALSE}
library(RPostgreSQL)

source('pw-file')
drv <- dbDriver("PostgreSQL")
ch <- dbConnect(drv, user= "postgres", password= pw, dbname= "graphs")

```

In case a password is needed, a file is `source`d to bring it in. Note that 
this block of code is not intended for execution in the database, so, 
in order to be able to exclude it from compilation into the PL/R function, 
the block is named with a prefix `prep`, for example, `prepConn`.

Next, we define the variable we want and execute the `pullDataSet` 
for our query as:

```{r prepLoad, eval= FALSE}
varClass <- "'m'"

dataSet <- pullDataSet(f= 'queryClass.sql'
                       , conn= ch
                       , varList= list(v_class = varClass))

dbDisconnect(ch)
```

This code block is named `prepLoad`.
If another type of data for the class of nodes exists, for example, 
temporal data, we add another call of the `pullDataSet` function for another 
query saved in a different .sql file.

Once the data is available in R, one may continue with any algorithm code 
chunks desired (as long as they are *not* named beginning with `prep`). 
Additionally, one may add code chunks which call 
visualization queries. These chunks are named with a prefix `vis` to 
distinguish them from the algorithm code.

Once the algorithm has completed, one must prepare a return data.frame to 
pass to the SQL environment. Typically, this is done by merging the final 
score data.frame with any interim data desired.

```{r plrOutput, eval= FALSE}
returnDf <- merge(scoreDf, interimDf)

```

One may add any other visualization code chunks after this point for 
unit-tests, sanity checks.

The final step is compiling the R markdown file into PL/R function 
definition, type, and execution queries. The `compileRToPLR` is used as 
follows:

```{r compile, eval= FALSE}
compileRToPLR(script= 'working-document.Rmd'
              , inputs= list(dat1= data1)
              , problemKey = 'class'
              , output= returnDf
              , name= 'schema.function'
              , outputFile= 'function.sql')
```

Note that one must name the data.frames which are inputs to the algorithm. 
The function will use the datatypes to define the PL/R function. The `class` 
variable is the one which defines the problem for a given R call, so the 
function needs to know that in order to define that as a scalar and not a 
vector. The output argument is the final data.frame expected at the end. 
By this time, the algorithm is complete and this data.frame should be 
available. Finally, the name of the PL/R function and the .sql file to write 
out are specified.
